A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-14 tex/codatadp.tex
more

To evaluate the decision procedure, we generated problems from existing
Isabelle formalizations using Sledgehammer \cite{paulson-blanchette-2010}.
Codatatypes being a recent addition to Isabelle
\cite{blanchette-et-al-2014-impl}, benchmarks are somewhat hard to come by. We
included the theory files from the Isabelle distribution and the \emph{Archive
of Formal Proofs} (AFP) \cite{klein-et-al-afp} that define codatatypes falling
within the supported fragment. We also included two unpublished theories by
Andreas Lochbihler and Peter Gammie. We complemented the benchmarks with some
theories about various list and tree data structures, to also test the
datatypes. The theory files were selected before running the experiments. Our
experimental data is publicly available \cite{our-eval-data}.

For each goal in each theory file, Sledgehammer was invoked to select
256~lemmas as axioms, which were then monomorphized and translated to SMT-LIB
along with the goal \cite{boehme-2012-phd}. The resulting problem was given to
CVC4, running for up to \textbf{TODO: XXX} seconds on the StarExec cluster
\cite{xxx}.

  * ran all problems several times, to try all four combinations of datatypes
    and codatatypes enabled or disabled -- also, for datatypes, we tried with
    and without the cyclicity rule; and for codatatypes, with and without the
    bisimilarity rule, so find out whether these somewhat expensive rules are
    useful in practice

  * lemmas/axioms often include basic properties of constructor and selectors,
    which are tautologies and ignored if (co)datatypes are enabled
  * allows a fair comparison: how much do (co)datatype support brings as
    opposed to just letting Sledgehammer do what it would normally do

  * experience with such features is not extremely high -- e.g. arithmetic,
    a most useful theory on lots of benchmarks, increases Sledgehammer's
    success rate by 2 percentage points, or 4\%, with Z3 in earlier work \cite{xxx}

  * so when analyzing statistics of this kind, with highly varied problems,
    we must not overestimate the importance of a single trick
  * on the other hand: every percentage point counts! cite Tom Hales

@@

\begin{table}[t!]
\begin{center}\begin{tabular}{l@{\kern1.5em}c@{\kern.5em}c@{\kern.5em}c@{\kern1.5em}c}
  & \HD{Distro} & \HD{AFP} & \HD{\landg} & All
\\[1pt] %%% TYPESETTING HACK
\midrule
%$\enatT$
\\[-9pt] %%% TYPESETTING HACK
No decision procedure
  & 800 & 922 & 450 & 2172 \\
Codatatypes without bisimilarity
  & 810 & 936 & 470 & 2216 \\
Codatatypes with bisimilarity
  & 811 & 939 & 472 & 2222 \\
Datatypes without cyclicity
  & 815 & 947 & 469 & 2231 \\
Datatypes with cyclicity
  & 818 & 953 & 470 & 2241 \\
Full decision procedure
  & 840 & 970 & 490 & 2300 \\[\jot]
Total number of goals
  & 1211\phantom{0} & 1121\phantom{0} & 899 & 3001
\end{tabular}\end{center}
\caption{Number of solved goals for the three benchmark suites}
\end{table}

* compare with BBP 2011!! --- potentially makes a more significant difference
  than arithmetic (although the benchmark suites are not directly comparable)

Do we lose any problems by enabling it?

* statistics, like how often do we get proofs that contain the cyclicity and
  bisimilarity rules

* and finally, we look in more detail at one or two such proofs, by presenting
  it and explaining it (and, before that, understanding it)

@@

This work is part of a larger program that aims at enriching automatic
provers with high-level features and at reducing the gap between automatic and
interactive theorem proving. As future work, we want to try out CVC4 and its
(co)datatypes for higher-order model finding in Isabelle, as an
alternative to the counterexample generator Nitpick
\cite{blanchette-nipkow-2010}. We also see some opportunities to enrich
SMT solvers with recursive and corecursive functions. Finally, it might be
possible to go further in terms of supporting nested and mixed (co)recursion
and quantified formulas over (co)datatypes in solvers.

>>>>

@@

\begin{table}[tbh!]
\begin{center}\begin{tabular}{l@{\kern1.5em}c@{\kern.5em}c@{\kern.5em}c@{\kern1.5em}c}
  & \HD{Distro} & \HD{AFP} & \HD{\landg} & All
\\[1pt] %%% TYPESETTING HACK
\midrule
%$\enatT$
\\[-9pt] %%% TYPESETTING HACK
No decision procedure
  & 800 & 922 & 450 & 2172 \\
Codatatypes without bisimilarity
  & 810 & 936 & 470 & 2216 \\
Codatatypes with bisimilarity
  & 811 & 939 & 472 & 2222 \\
Datatypes without cyclicity
  & 815 & 947 & 469 & 2231 \\
Datatypes with cyclicity
  & 818 & 953 & 470 & 2241 \\
Full decision procedure
  & 840 & 970 & 490 & 2300 \\[\jot]
Total number of goals
  & 1211\phantom{0} & 1121\phantom{0} & 899 & 3001
\end{tabular}\end{center}
\caption{Number of solved goals for the three benchmark suites}
\label{tab:bench}
\end{table}

To evaluate the decision procedure, we generated problems from existing
Isabelle formalizations using Sledgehammer \cite{paulson-blanchette-2010}.
Codatatypes being a recent addition to Isabelle
\cite{blanchette-et-al-2014-impl}, benchmarks are somewhat hard to come by. We
included the theory files from the Isabelle distribution (the Distro) and the \emph{Archive
of Formal Proofs} (AFP) \cite{klein-et-al-afp} that define codatatypes falling
within the supported fragment. We also included two unpublished theories by
Andreas Lochbihler and Peter Gammie (\landg). We complemented the benchmarks
with some theories about various list and tree data structures, to also test
the datatypes.
The theory files were selected before running the experiments.
Our experimental data is publicly available \cite{our-eval-data}.

For each goal in each theory file, Sledgehammer was invoked to select
256~lemmas as axioms, which were then monomorphized and translated to SMT-LIB
along with the goal \cite{boehme-2012-phd}. The resulting problem was given to
CVC4, running for up to \textbf{TODO: XXX} seconds on the StarExec cluster
\cite{xxx}. Due to the lack of machinery for reconstructing inferences about
(co)datatypes in Isabelle, CVC4 is trusted as an oracle in these experiments.

CVC4 was run on each problem several times, with the support for datatypes and
codatatypes either enabled or disabled. The contributions of the cyclicity and
bisimilarity rules were also measured, so find out whether these somewhat
expensive rules are useful in practice. Even when the decision procedure is
disabled, some of the generated axioms typically include basic properties of
constructors and selectors, which the decision procedure would recognize as
tautologies. This allows us to answer the question, \relax{What are the
benefits of activating the decision procedure as opposed to letting
Sledgehammer do what it would normally do?}

\newcommand\BAD[1]{\textcolor{red}{\textbf{#1}}}

%  * experience with such features is not extremely high -- e.g. arithmetic,
%    a most useful theory on lots of benchmarks, increases Sledgehammer's
%    success rate by 2 percentage points, or 4\%, with Z3 in earlier work \cite{xxx}

%  * so when analyzing statistics of this kind, with highly varied problems,
%    we must not overestimate the importance of a single trick
%  * on the other hand: every percentage point counts! cite Tom Hales

The results are summarized in Table~\ref{tab:bench}. The decision procedure
makes a difference across all theories, and an overall difference of
\BAD{4.7\%}. Moreover, every aspect of the procedure, including the more
expensive rules, make a contribution. What is not visible in the table, but
readily available when looking at the raw data \cite{our-eval-data}, is that
the stronger decision procedures subsume the weaker ones. In the context of
Sledgehammer, the power of interpreted (co)datatype is comparable with that of
arithmetic \cite{blanchette-et-al-2013-smt}.

%* and finally, we look in more detail at one or two such proofs, by presenting
%  it and explaining it (and, before that, understanding it)

@@

This work is part of a larger program that aims at enriching automatic provers
with high-level features and at reducing the gap between automatic and
interactive theorem proving. As future work, we want to implement proof
reconstruction for (co)datatype inferences in Isabelle. We also want to try
out CVC4 and its (co)datatypes for higher-order model finding in Isabelle, as
an alternative to the counterexample generator Nitpick
\cite{blanchette-nipkow-2010}. We also see some opportunities to enrich
SMT solvers with recursive and corecursive functions. Finally, it might be
possible to go further in terms of supporting nested and mixed (co)recursion
and quantified formulas over (co)datatypes in solvers.
